Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)

### 2022-11-18 analysis/ms_november/hrs_2022-11-07.Rmd
2022-11-18

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    correspondingauthor: true
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "the footnote"
abstract: |
  This is the abstract.

  It consists of two paragraphs.
#  We use historical forest harvest data from Alberta, Canada to develop
# a description of the spatial and temporal distributions of the
# availability of forest harvest residues for the period 1990---2015.
# Using a GIS coverage of the road network in Alberta, we developed
# estimates of round-trip haul costs from each harvest location to each of
# three potential bioenergy plant locations. This information was used to
# create supply curves (or cost-availability curves) for each of the three
# potential locations and each of the 26 years. This information is used
# to develop descriptions of the variability in the delivered costs of
# forest residue for bioenergy use. These descriptions of variability are
# historical, but could be useful for potential investors in the bioenergy sector.
keywords:
  - argh
  - billy
journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

@@

Humans have used biofuels since learning to control fire a million or so
years ago. In developed economies, early biofuels (*e.g*. wood,
charcoal, dung, peat) were largely displaced by coal during the
industrial revolution, and by petroleum starting in the mid-nineteenth
century. In recent times, biofuels are once again becoming an important
source of energy because of their potential to reduce greenhouse gas
emissions. In the European Union, the United Kingdom, and Japan, many
heat and power plants have been converted to use wood pellets instead of
coal or natural gas.

@@

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
waste residue left over from food or forest products production. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that will use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (2022-07-08), the plant is not yet operational.

Technologies to pelletize wood are more established than technologies to
produce liquid fuels from wood. It was projected that Canada would
produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill residues.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest: this
practice is well-established in Europe.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

A number of studies have shown that collection and transportation
costs may be the biggest impediment to widespread use of forest
residues in bioenergy systemsX
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depends on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (@Caputo2005 @Leistritz2007 cited in @Stephen2010),a number of studies have shown that
collection and transportation costs may be the biggest impediment to
widespread use of forest residues in bioenergy systems. reliable feedstock supplies.
Since a
forest-residue based bioenergy plant is dependent on stemwood harvesting
for feedstock, any variability in harvesting directly influences the
volume of residues that would be available. Therefore, considering the
long-term variability associated with feedstocks in different potential
locations could be important in choosing the optimal site for locating a
bioenergy plant.

and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Past studies regarding bioenergy plant locations have
improved our understanding of feedstock availability and constraint.
One key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant.

Although few studies consider long-term forest residue variability for
the purposes of optimizing the location of a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility (Chen and Fay
2011) Chen2012, types of storage infrastructure for bioenergy plant
feedstocks Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from three potential locations. More
specifically, we use precise information on the location and year of
historical timber harvest in the western Canadian province of Alberta to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed from the road network in the province.
The spatial and temporal variability of forest residues results in
variability in the transport cycle time from the residue collection
areas to the bioenergy plant, which in turn leads to variability in
delivered residue cost. From this information we develop historical cost
curves for collecting and delivering forest residues to each of
the three potential bioenergy plant locations for each of 26 years in
our study period (1990--2015). When considering locations for the
potential bioenergy plants, we conduct sensitivity analysis on 3
different bioenergy plant capacities, as defined by annual forest
residue feedstock requirements: 200 Gg/a corresponding to a small wood
pellet mill, 400 Gg/a corresponding to a large pellet mill or a small
biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops supply curves (cost-availability curves) based on
a road network and historical harvests. However, their take on temporal
variability is limited to comparing two 5-year periods. As well, they
limit their analyses to the boundaries of timber supply areas. The only
boundary that limits our analysis is the provincial border between
Alberta and British Columbia.

Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop supply curves at national and regional
levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost. Instead, we consider the
spatial location of all the harvest sites over a 26-year period
(1990-2015) and calculate the delivery cost for each harvest site in the
year it was harvested.

In the next section we describe our study area and data sources. We then
describe how we aggregate and analyze the data. We present our
distributional results in the form of marginal delivered biomass cost
for each of the 26 years in the study period, and for each of the three
alternative plant locations. We conclude with a discussion of financial
considerations regarding forest residues as a biofuel feedstock, with
relevance to forest companies, policy makers, and bioenergy investors.

@@

We use the province of Alberta in western Canada as our study area. It
occupies 661 848 km^2^. Its southern boundary is 49°N latitude, northern
boundary is 60°N latitude, eastern boundary is 110°W longitude, and its
western boundary is defined by 120°W longitude and the Great Divide of
the Rocky Mountains (Figure 1).

[ Figure 1 about here ]

@@

The Alberta government requires FMA holders to develop a forest
management plan for the FMAA. The forest management plans were used to
obtain information on utilization standards and average piece size of
harvested trees for the FMAA @Alberta2021.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator @NFIS2021b, described in
section 2.2.3.

@@

Most of the effort expended on this study was spent in aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with RStudio @RStudioTeam2021, and tidyverse @RStudioTeam2021.

We develop a model based on the assumption that forest residue
extractors follow a greedy optimization procedure: they take the
cheapest (closest) residue first, and stop when they have enough to
satisfy the annual biomass requirements for the three scales of
bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
residue yield (Mg ha^-1^), cycle times (h), and harvest areas (ha) are
the three essential summary parameters used by the greedy optimization
model.

@@

[ Figure 2 about here ]

@@

We assume that the residue will be transported from the forest using a
live floor chip van with 100 m^3^ capacity. The USDA Forest Service's
Forest Residue Transportation Costing Model @Rummer2005 assumes a default
solid volume factor of 0.3 for biomass meaning that a 100 m^3^ capacity
trailer could hold the equivalent of 30 m^3^ of solid biomass. This
equates to 11.40 Mg of biomass using the specific gravity of lodgepole
pine wood and bark on a green volume basis (0.38) (Ref. @Miles2009).

Timber Tracks produces a publication describing forestry equipment
hourly rates @TT2020. The rate reported for a tandem tractor with a
tandem trailer was 157.31 \$·h^-1^. We assume that the tractor-trailer
combination used to haul biomass would cost a similar amount.

Therefore, the cost of transporting residual biomass was set to 13.80
\$·Mg^-1^·h^-1^.

@@

. The information collected above was collated into a data set which had
a record for each combination of township, plant location, year, and
capacity. The fields in the data set were of township, plant location,
year, and capacity, cycle time, harvest area, and residue yield. This
summary data set is the input to the greedy optimization model.

@@

The model uses a greedy algorithm whereby the lowest cost residue for a
location is extracted first. The model is summarized using pseudocode in
Figure ?.

@@

## Distributional results

Our results assume that only one of the three potential bioenergy plants
would be built. If more than one plant was to be built, the residue
collection areas would overlap, thereby reducing the available residue
for the second plant. Figures 5 and 6 display marginal cycle time (h)
and marginal costs (\$·Mg^-1^) for each of the 26 years in the study
period, and for the 3 potential plant locations. Curves for 1990, 2009,
and 2005 are highlighted represented the years of minimum, median, and
maximum provincial harvest volume in the period. The inter-year
variation in marginal costs increases as the cumulative amount of
residue collected increases. We can also see that the relative position
of the years in terms of cost varies between the plant locations:
*e.g.,* the year 2005 was the year of highest provincial harvest, but
from a residue availability standpoint was not as good for Hinton as it
was for Edson and Whitecourt.

[ Figure 5 about here ]

[ Figure 6 about here ]

@@

[ Figure 7 about here ]

Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
plant capacity, and year. Overall, the average costs in Edson and
Whitecourt are quite similar. Hinton is usually more expensive across
years and capacities. The differences become clearer at a capacity of
800 Gg/a, for which Whitecourt is the lowest cost location for most
years.

[ Figure 8 about here ]

@@

\[ Table 2 about here ]

@@

[ Figure 9 about here ]

Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
medium, and low harvest years, respectively) for Whitecourt. The area
required to supply the bioenergy plant increases greatly in low timber
harvest years, particularly for the 800 Gg/a capacity. Similar maps
have been produced for the other locations and all 26 years and are
available at the University of Alberta's Education and Research Archive
(ERA). **cite** Notice that the supply areas include areas outside the
Green Area with very little harvest. The contours indicate the cycle
time associated with each of the capacities. Because of the cost of
moving equipment such as the grinder and loader, it is unlikely that the
small, dispersed harvested areas would be accessed, but the volumes are
small enough that we ignore them. We have created animations of these
contour maps for all 26 years in the study period for the 3 locations
(available at ERA). These provide another way of visualizing the
inter-annual variation.

[ Figure 10 about here ]

Whitecourt is lowest cost location for most years at all capacities.
There is a noticeable spike in costs 1999, which may be related to
salvage harvest following the 1998 Virginia Hills wildfire \[56\]. Note
that this spike also affects Hinton and Edson at 800 Gg/a capacity,
because they reach into the Whitecourt area. The average cost of
delivered forest residues are lower for the lower capacity plants, as
they do not need to collect residues from as far to meet their biomass
input requirements. In other words, the average cycle time is lower for
the lower capacity bioenergy plants, which results in a lower average
delivered residue cost.

@@

*** Which areas are best for locating a potential bioenergy plant when
considering residue collection and transport costs? *** We focus on the
area surrounding the towns of Hinton, Edson, and Whitecourt in
west-central Alberta. These three potential bioenergy plant locations
were chosen as they host several forest products companies and are
located near forest areas with substantial harvesting activity. For each
location, we conduct sensitivity analysis on three alternative bioenergy
plant capacities as measured by residue feedstock requirements: 200
Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
capacity the three locations are almost indistinguishable in terms of
average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
is important to note that the average cost of supplying an 800 Gg/a
plant is over 40% greater than supplying a 200 Gg/a plant.

*** How much residue volume is spread over what area? *** The residue
supply areas for the Whitecourt location shown in Figure 7 suggest that
if a 400 or 800 Gg/a bioenergy plant was to be constructed there
would only be enough available residue to construct one plant in our
study area. Whereas if a 200 Gg/a plant was deemed more viable, our
findings suggest that two plants could be constructed in the study area
-- one in Whitecourt and one in Hinton. This makes for an interesting
investment choice, since a single 800 Gg/a bioenergy plant would
likely face an average delivered residue cost that is over 40% higher
than the cost for the 200 Gg/a plants. Recall from above that 200
Gg/a is the appropriate scale for a pellet plant and 800 Gg/a
is more representative of a biorefinery. Given that dried and compacted
pellets are cheaper to transport than moist and bulky residues, it could
be more viable to first transport residues to pellet plants, and then
transport the pellets (perhaps via rail) from multiple pellet plants to
a biorefinery. The optimal result of this analysis will depend on the
cost of pellet production relative to the cost savings (transportation
and production) to the biorefinery from using pellets instead of
residues. Although outside the scope of this work, we plan to conduct
this analysis in the future.

*** Would residue volumes be consistently available on a year-to-year
basis?***  Given that a forest residue-based bioenergy plant is dependent
on forest harvesting, transportation costs increase (trucks must travel
further to meet the feedstock capacity requirement) in years when
harvesting levels are low or harvest areas are further from the
bioenergy plant. Our results showed considerable variability in both
marginal and average residue extraction costs from year-to-year for the
400 and 800 Gg/a capacities, suggesting that residue cost
variability is an important source of risk for medium and large
bioenergy plants. Our finding that the cost variability is lower for the
200 Gg/a might once again suggest that a 200 Gg/a is a more
appropriate scale for a forest residue-based bioenergy plant in Alberta.

@@

This work started as an extension to Irene Onyango's Master of
Agriculture research.

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    correspondingauthor: true
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
  - name: M.K. Luckert
    email: mluckert@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1

abstract: |
  We use historical forest harvest data from Alberta, Canada to develop
  a description of the spatial and temporal distributions of the
  availability of forest harvest residues for the period 1990---2015.
  Using a GIS coverage of the road network in Alberta, we developed
  estimates of round-trip haul costs from each harvest location to each of
  three potential bioenergy plant locations. This information was used to
  create supply curves (or cost-availability curves) for each of the three
  potential locations and each of the 26 years. This information is used
  to develop descriptions of the variability in the delivered costs of
  forest residue for bioenergy use. These descriptions of variability are
  historical, but could be useful for potential investors in the bioenergy sector.
keywords:
  - residue
  - road network
  - supply curve
journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: historicalResidueSupply.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-vancouver
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: default
---

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis/ms_november")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(rlang)
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

@@

Humans have used biomass as fuel since learning to control fire a million or so
years ago. In developed economies, early biofuels (*e.g*., wood,
charcoal, dung, peat) were largely displaced by coal during the
industrial revolution, and by petroleum starting in the mid-nineteenth
century. In recent times, biofuels are once again becoming an important
source of energy because of their potential to reduce greenhouse gas
emissions. In the European Union, the United Kingdom, and Japan, many
heat and power plants have been converted to use wood pellets instead of
coal or natural gas.

@@

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
waste residue left over from food or forest products production. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that will use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (2022-11-07), the plant is not yet operational.

Technologies to pelletize wood are more established than technologies to
produce liquid fuels from wood. It was projected that Canada would
produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
exported, largely to the United Kingdom, Japan, and Europe @Watters2021. Most
of the wood pellet production in Canada comes from forest product mill
residues, such as sawdust and bark. There is interest, however, in
expanding the biomass source to the forest by collecting the residues
left within harvested areas: a practice that is rare in Canada, but
well-established in Europe.

Forest residues are generated as by-products of conventional forest
harvesting operations and include the treetops, branches, and
non-merchantable logs. In Canada, these residues are typically left
behind in slash piles, which are burned in winter to mitigate against
the risk of wildfires. Not all forest residues are available for
extraction, however, and estimates of residue availability should
consider ecological sustainability and technical accessibility. At a
national scale, a previous study by Yemshanov *et al.* @Yemshanov2018 estimated the
annual supply of harvestable residual biomass in Canada to be between
16.5 and 20.0 Tg/a in scenarios that included both ecological and
technical accessibility limitations. In this paper we express biomass
using SI units, and all quantities of biomass will be expressed on an
oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.

Despite Canada's huge potential supply of bioenergy feedstock, there are
key questions about the financial viability of forest-residue based
bioenergy investments: How much volume is spread over what area? Which
areas are best for locating a potential bioenergy plant when considering
forest residue collection and transport costs? Would forest residue
volumes be consistently available on a year-to-year basis? We attempt to
answer these questions in this paper.

There are several studies regarding plant location decisions for forest
residue-based bioenergy plants, as reviewed by Johnson *et al*. @Johnson2012.
Although these studies have improved our understanding of feedstock
availability, they tend to use limited datasets and often focus on
average levels of residue availability. A key area that has received
little attention in the bioenergy plant location literature is
consideration of feedstock supply variability over longer periods of
time. Instead of just focusing on average residue levels, a bioenergy
investor is likely to be concerned about the financial impact of
variable annual residue availability , which is the focus of our study.

Availability of forest residues for a bioenergy plant largely depends on
commercial harvesting operations and the production of traditional
forest products. Fluctuations in the volume of stemwood harvested for
forest products has a direct impact on residues available for biofuels.
Factors such as the 2008 housing crisis and the ongoing softwood lumber
dispute between Canada and the United States of America have resulted in
sawmill closures and reductions in forest harvesting activities in
Canada @deAvillez2014 @Spelter2009. This relationship between variable harvest
volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue supply in Alberta
as a function of lumber prices.

Biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (from (@Caputo2005; @Leistritz2007)  who are
cited in @Stephen2010). A number of studies have shown that collection and
transportation costs may be the biggest impediment to widespread use of
forest residues in bioenergy systems (*e.g.* @Aulakh2008). Since a forest-residue based
bioenergy plant is dependent on stemwood harvesting for feedstock, any
variability in harvesting directly influences the volume of residues
that would be available. Therefore, considering the long-term
variability associated with feedstocks in different potential locations
would be important in choosing the optimal site for locating a bioenergy
plant.

Although few studies consider long-term forest residue variability for
the purposes of optimizing the location of a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility  @Chen2012, types of storage infrastructure for bioenergy plant
feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from alternative potential locations.
More specifically, we use precise information on the location and year
of historical timber harvest in the province of Alberta, Canada to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed from a digital road network for the
province. The spatial and temporal variability of forest residues
results in variability in the transport cycle time from the residue
collection areas to the bioenergy plant, which in turn leads to
variability in delivered residue cost. From this information we develop
historical cost-availability curves for collecting and delivering
forest residues to each of the three potential bioenergy plant locations
for each of 26 years in our study period (1990--2015). When considering
locations for the potential bioenergy plants, we conduct sensitivity
analysis on 3 different bioenergy plant capacities, as defined by annual
forest residue feedstock requirements: 200 Gg/a corresponding to a small
wood pellet mill, 400 Gg/a corresponding to a large pellet mill or a
small biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops cost-availability curves based on a road network
and historical harvests. However, their consideration of temporal
variability is limited to comparing two 5-year periods. As well,
FPInnovations limit their analyses to the boundaries of timber supply
areas. The only boundary considered our analysis is the provincial
boundary of Alberta.

Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop cost-availability curves at national and
regional levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost.

In the next section we describe our study area and data sources. We then
describe how we aggregate and analyze the data. We present our results
in the form of marginal delivered biomass cost for each of the 26 years
in the study period, and for each of the three alternative plant
locations. We conclude with a discussion of financial considerations
regarding forest residues as a biofuel feedstock, with relevance to
forest companies, policy makers, and bioenergy investors.

@@

We use the province of Alberta, Canada as our study area. It
occupies 661 848 km^2^. Its southern boundary is 49°N latitude, northern
boundary is 60°N latitude, eastern boundary is 110°W longitude, and its
western boundary is defined by 120°W longitude and the Great Divide of
the Rocky Mountains (Figure 1).

` `  
` `  
[ Figure 1 about here ]  
` `  
` `

@@

The Alberta government requires FMA holders to develop a forest
management plan for the Forest Management Agreement Area (FMAA). The
forest management plans were used to obtain information on utilization
standards and average piece size of harvested trees for the FMAA @Alberta2021.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator@NFIS2021b.

@@

Most of the effort expended on this study went into aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with Rstudio @RStudioTeam2021, and tidyverse @Wickham2019.

We develop a model based on the assumption that forest residue
extractors follow a greedy optimization procedure: they take the
cheapest (closest) residue first, and stop when they have enough to
satisfy the annual biomass requirements for the three scales of
bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
residue yield (Mg ha^-1^), cycle times (h), and harvest areas (ha) are
the three essential summary parameters used by the greedy optimization
mode

@@

` `  
` `  
[ Figure 2 about here]
` `  
` `

@@

We assume that the residue will be transported from the forest using a
live floor chip van with 100 m^3^ capacity. The USDA Forest Service's
Forest Residue Transportation Costing Model @Rummer2005 assumes a default
solid volume factor of 0.3 for biomass meaning that a 100 m^3^ capacity
trailer could hold the equivalent of 30 m^3^ of solid biomass. This
equates to 11.40 Mg of biomass using the specific gravity of lodgepole
pine wood and bark on a green volume basis (0.38) @Miles2009.

Timber Tracks produces a publication describing forestry equipment
hourly rates @TT2020. The rate reported for a tandem tractor with a
tandem trailer was 157.31 \$·hr^-1^. We assume that the tractor-trailer
combination used to haul biomass would cost a similar amount.

Therefore, the cost of transporting residual biomass was set to 13.80
\$·Mg^-1^·hr^-1^.

@@

The information collected above was collated into a data set which had a
record for each combination of township, plant location, year, and
capacity. The fields in the data set were of township, plant location,
year, and capacity, cycle time, harvest area, and residue yield. This
summary data set is the input to the greedy optimization model

@@

The model uses a greedy algorithm whereby the lowest cost residue for a
location is extracted first. The model is summarized using pseudocode in
Figure 4.

` `  
` `  
[ Figure 4 about here ]
` `  
` `

@@

Our results assume that only one of the three potential bioenergy plants
would be built. If more than one plant was to be built, the residue
collection areas would overlap, thereby reducing the available residue
for the second plant. We will consider simultaneous location of more
than one plant in a subsequent study. Figures 5 and 6 display marginal
cycle time (h) and marginal costs (\$·Mg^-1^) for each of the 26 years
in the study period, and for the 3 potential plant locations. Curves for
1990, 2009, and 2005 are highlighted represented the years of minimum,
median, and maximum provincial harvest volume in the period. The
inter-year variation in marginal costs increases as the cumulative
amount of residue collected increases. We can also see that the relative
position of the years in terms of cost varies between the plant
locations: *e.g.,* the year 2005 was the year of highest provincial
harvest, but from a residue availability standpoint was not as good for
Hinton as it was for Edson and Whitecourt.

` `  
` `  
[ Figure 5 about here]
` `  
` `

` `  
` `  
[ Figure 6 about here]
` `  
` `

@@

` `  
` `  
[Figure 7 about here ]
` `  
` `

Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
plant capacity, and year. Overall, the average costs in Edson and
Whitecourt are quite similar. Hinton is usually more expensive across
years and capacities. The differences become clearer at a capacity of
800 Gg·yr^-1^, for which Whitecourt is the lowest cost location for most
years.

` `  
` `  
[Figure 8 about here ]
` `  
` `

@@

` `  
` `  
[ Table 2 about here ]
` `  
` `

@@

` `  
` `  
[ Figure 9 about here ]
` `  
` `

Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
medium, and low harvest years, respectively) for Whitecourt. The area
required to supply the bioenergy plant increases greatly in low timber
harvest years, particularly for the 800 Gg·yr^-1^ capacity. Similar maps
have been produced for the other locations and all 26 years and are
available at the University of Alberta's Education and Research Archive
(ERA). **will cite when location is finalized** Notice that the supply areas include areas outside the
Green Area with very little harvest. The contours indicate the cycle
time associated with each of the capacities. Because of the cost of
moving equipment such as the grinder and loader, it is unlikely that the
small, dispersed harvested areas would be accessed, but the volumes are
small enough that we ignore them. We have created animations of these
contour maps for all 26 years in the study period for the 3 locations
(available at ERA). These provide another way of visualizing the
inter-annual variation.

` `  
` `  
[ Figure 10 about here ]
` `  
` `

Whitecourt is lowest cost location for most years at all capacities.
There is a noticeable spike in costs 1999, which may be related to
salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
that this spike also affects Hinton and Edson at 800 Gg·yr^-1^ capacity,
because they reach into the Whitecourt area. The average cost of
delivered forest residues are lower for the lower capacity plants, as
they do not need to collect residues from as far to meet their biomass
input requirements. In other words, the average cycle time is lower for
the lower capacity bioenergy plants, which results in a lower average
delivered residue cost.

@@

**Which areas are best for locating a potential bioenergy plant when
considering residue collection and transport costs?** We focus on the
area surrounding the towns of Hinton, Edson, and Whitecourt in
west-central Alberta. These three potential bioenergy plant locations
were chosen as they host several forest products companies and are
located near forest areas with substantial harvesting activity. For each
location, we conduct sensitivity analysis on three alternative bioenergy
plant capacities as measured by residue feedstock requirements: 200
Gg·yr^-1^, 400 Gg·yr^-1^, and 800 Gg·yr^-1^. For the 200 Gg·yr^-1^
capacity the three locations are almost indistinguishable in terms of
average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg·yr^-1^,
Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg·yr^-1^,
Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
is important to note that the average cost of supplying an 800 Gg·yr^-1^
plant is over 40% greater than supplying a 200 Gg·yr^-1^ plant.

**How much residue volume is spread over what area?** The residue
supply areas for the Whitecourt location shown in Figure 7 suggest that
if a 400 or 800 Gg·yr^-1^ bioenergy plant was to be constructed there
would only be enough available residue to construct one plant in our
study area. Whereas if a 200 Gg·yr^-1^ plant was deemed more viable, it
is possible that two plants could be constructed in the study area --
one in Whitecourt and one in Hinton. However, we will leave this
question to a subsequent study.

This makes for an interesting investment choice, since a single 800
Gg·yr^-1^ bioenergy plant would likely face an average delivered residue
cost that is over 40% higher than the cost for the 200 Gg·yr^-1^ plants.
Recall from above that 200 Gg·yr^-1^ is intended to represent a pellet
plant and 800 Gg·yr^-1^ is more representative of a biorefinery. Given
that dried and compacted pellets are cheaper to transport than moist and
bulky residues, it could be more viable to first transport residues to
pellet plants, and then transport the pellets from multiple pellet
plants to a biorefinery. The optimal result of this analysis will depend
on the cost of pellet production relative to the cost savings
(transportation and production) to the biorefinery from using pellets
instead of residues. Although outside the scope of this work, we plan to
conduct this analysis in the future.

**Would residue volumes be consistently available on a year-to-year
basis?** Given that a forest residue-based bioenergy plant is dependent
on forest harvesting, transportation costs increase (trucks must travel
further to meet the feedstock capacity requirement) in years when
harvesting levels are low or harvest areas are further from the
bioenergy plant. Our results showed considerable variability in both
marginal and average residue extraction costs from year-to-year for the
400 and 800 Gg·yr^-1^ capacities, suggesting that residue cost
variability is an important source of risk for medium and large
bioenergy plants. Our finding that the cost variability is lower for the
200 Gg·yr^-1^ might once again suggest that a 200 Gg·yr^-1^ is a more
appropriate scale for a forest residue-based bioenergy plant in Alberta.

@@

This work began as an extension to Irene Onyango's Master of
Agriculture research.
